Modelling the emergence of whisker barrels

### 2020-06-16 paper/briefpaper/brief.tex
plodding through with edits

\documentclass[9pt,lineno]{elife}

@@

% Some document-defined commands
\newcommand*\dif{\mathop{}\!\mathrm{d}}
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\newcommand{\code}[1]{\textsf{#1}}

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \cmnt{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}) The second term on the
right quantifies the coupling between axon branching and synaptogenesis. Here
  $\chi_i=0$ is a placeholder.

@@

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}. \textbf{B} The strengths of interaction
      $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
      projections by the lengths of green and blue arrows respectively, assuming
      that similar fields aligned to the posterior-anterior and medial-lateral
      axes in the ventroposterior medial nucleus of the thalamus are sampled at
      the locations of putative barreloid centers (reconstructed from
      \citealp{haidarliu_size_2001}, \cmnt{Fig.\,5b}). \textbf{C} Simulation results for parameters
      \cmnt{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
      $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours indicate the thalamic
      projection for which the connection density is maximal and black lines
      delineate boundaries (see Movie
      S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
      simulation approaching that obtained from \cmnt{real} barrels in \textbf{A} (dotted
      line); black squares \cmnt{give a} measure \cmnt{of} the correspondence
      between the real and
      simulated barrel shapes, $\eta$ (the sum of differences in
      area; units \cmnt{mm$^2$});
      \cmnt{grey hexagons show the measure of localization, $\omega =
        \oiint_{S} \left( \mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i\right) \mathrm{d}S$. }
      \textbf{E} \cmnt{The localization measure plotted across the domain
        shows that the structure in the connection density which has developed
        by $t=30000$ is largely absent at $t=1000$. The
        colour map
        indicates the values of  $\mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i$. Overlaid
        contours show $c_i > 0.95 \mathrm{max}_i(c_i)$.} All scale bars 1\,mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\cmnt{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{blue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}
  \color{black}
\end{equation}
%
\cmnt{which contributes to the \emph{flux of axonal branching} as an
  additional diffusion-like term which makes the branching density for a given
  projection diffuse away from regions} where the branches of other projections
become dense. \cmnt{$\epsilon$ parameterizes the strength of this competitive
  interaction.} Note that this operation is local to individual afferent
projections.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. \cmnt{Thus, we suggest that there exist a
  pair of signalling molecules whose expression varies linearly (with
  orthogonal gradients) across the thalamus, analgous to the ephrins, whose
  existence is well known in the cortex. Thalamocortical axons growing from a
  region in the thalamus with high expression of one of these molecules will
  interact with a corresponding cortical signalling molecule more strongly
  than an axon emanating from a region with low expression of the thalamic
  signal.}  \cmnt{We defined two orthogonal, linear gradients oriented in the
  anterior-posterior and medial-lateral directions and used the coordinates of
  41 barreloid centers recreated from \cite{haidarliu_size_2001},
  corresponding to the macrovibrissae, to populate 41 pairs of $\gamma$
  values.}  (Fig.\,\ref{fig:main}B). \cmnt{In other words, we simply scaled
  the barreloid coordinates in an anterior-posterior/medial-lateral coordinate
  system to create the $\gamma$ values in the range [-2,2]. Note that positive
  \emph{and} negative $\gamma$ values were generated; the origin of the
  coordinate system was located at the approximate center of the thalamic
  barreloid field. This prevented an overall drift of the cortical pattern
  away from the centroid of the barrel field and may explain why
  coordinate-giving transcription factors and the ephrins are often found as
  pairs acting in opposing directions.}  \cmnt{We concentrated on
  macrovibrissae (excluding microvibrissae) because it is easy to identify
  their corresponding structures both in the thalamus and in the cortex. We
  also excluded the 5 rhinal whiskers because their cortical representation is
  in a separate island, distinct, and somewhat removed from the representation
  of the macrovibirissae.}

>>>>

\documentclass[9pt,lineno,draft]{elife}

@@

% Some document-defined commands
\newcommand*\dif{\mathop{}\!\mathrm{d}}
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\stu}[1]{\textcolor{red}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\newcommand{\code}[1]{\textsf{#1}}

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \cmnt{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}). The second term on the
right quantifies the coupling between axon branching and synaptogenesis. Here
  $\chi_i=0$ is a placeholder.

@@

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}. \textbf{B} The strengths of interaction
      $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
      projections by the lengths of green and blue arrows respectively, assuming
      that similar fields aligned to the posterior-anterior and medial-lateral
      axes in the ventroposterior medial nucleus of the thalamus are sampled at
      the locations of putative barreloid centers (reconstructed from
      \citealp{haidarliu_size_2001}, \cmnt{Fig.\,5b}). \textbf{C} Simulation results for parameters
      \cmnt{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
      $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours indicate the thalamic
      projection for which the connection density is maximal and black lines
      delineate boundaries (see Movie
      S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from \stu{the}
      simulation approaching that obtained from \cmnt{real} barrels in \textbf{A} (dotted
      line); black squares \stu{provide a} measure \cmnt{of} the correspondence
      between the real and simulated barrel shapes, $\eta$ (the sum of differences in
      area; units \cmnt{mm$^2$});
      \cmnt{grey hexagons \stu{provide a measure of how selectively each cortical site is innervated}, $\omega = \oiint_{S} \left( \mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i\right) \mathrm{d}S$. }
      \textbf{E} \stu{Plotted across the cortical sheet, the selectivity develops to reveal an alignment with the emergent barrel boundary shapes}
      \cmnt{The colour map indicates the values of  $\mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i$. Overlaid
        contours show $c_i > 0.95 \mathrm{max}_i(c_i)$.} All scale bars 1\,mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\cmnt{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{blue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\stu{This term contributes to the \emph{flux of axonal branching} as an
 additional source of diffusion, which reduces the branching density for a given projection where the branches of other projections are dense.} \stu{The strength of this additional competition is parameterized by $\epsilon$}. Note that this operation is local to individual afferent
projections.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. \stu{As such, we consider linear gradients aligned to anterior-posterior and medial-lateral axes through the VPM, and as $\gamma$-values use the coordinates of the barreloid centers in these axes as they appear in \citep{haidarliu_size_2001} (Fig.\,\ref{fig:main}B).} \stu{Complementary thalamic and cortical molecular guidance gradients of this form have been well established experimentally \citep{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}, e.g., cells originating in the medial part of ventrobasal thalamus (comprising VPM and VPL) express high levels of EphA receptors and project to the lateral part of S1, which expresses low levels of ephrin-A5, and cells originating in the lateral part of ventrobasal thalamus express low levels of EphA receptors and project to the medial part of S1, which expresses high levels of ephrin-A5 (see Gao, 1998; Dufour et al., 2003; Vanderhaegen \& Polleux, 2004; Speer \& Chapman, 2005)}

%\cmnt{In other words, we simply scaled the barreloid coordinates in an anterior-posterior/medial-lateral coordinate system to create the $\gamma$ values in the range [-2,2]. Note that positive \emph{and} negative $\gamma$ values were generated; the origin of the coordinate system was located at the approximate center of the thalamic barreloid field. This prevented an overall drift of the cortical pattern away from the centroid of the barrel field and may explain why coordinate-giving transcription factors and the ephrins are often found as pairs acting in opposing directions.}  \cmnt{We concentrated on macrovibrissae (excluding microvibrissae) because it is easy to identify their corresponding structures both in the thalamus and in the cortex. We also excluded the 5 rhinal whiskers because their cortical representation is in a separate island, distinct, and somewhat removed from the representation of the macrovibirissae.}}

%\cmnt{Thus, we suggest that there exist a pair of signalling molecules whose expression varies linearly (with orthogonal gradients) across the thalamus, analgous to the ephrins, whose existence is well known in the cortex. Thalamocortical axons growing from a region in the thalamus with high expression of one of these molecules will interact with a corresponding cortical signalling molecule more strongly than an axon emanating from a region with low expression of the thalamic signal.}

%\cmnt{We defined two orthogonal, linear gradients oriented in the anterior-posterior and medial-lateral directions and used the coordinates of 41 barreloid centers recreated from \cite{haidarliu_size_2001}, corresponding to the macrovibrissae, to populate 41 pairs of $\gamma$ values.}  (Fig.\,\ref{fig:main}B). \cmnt{In other words, we simply scaled the barreloid coordinates in an anterior-posterior/medial-lateral coordinate system to create the $\gamma$ values in the range [-2,2]. Note that positive \emph{and} negative $\gamma$ values were generated; the origin of the coordinate system was located at the approximate center of the thalamic barreloid field. This prevented an overall drift of the cortical pattern away from the centroid of the barrel field and may explain why coordinate-giving transcription factors and the ephrins are often found as pairs acting in opposing directions.}  \cmnt{We concentrated on macrovibrissae (excluding microvibrissae) because it is easy to identify their corresponding structures both in the thalamus and in the cortex. We also excluded the 5 rhinal whiskers because their cortical representation is in a separate island, distinct, and somewhat removed from the representation of the macrovibirissae.}
